---
title: "Rathmell_Dori_A02_CodingBasics.Rmd"
author: "Dori Rathmell"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: inline
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics Day 1

1.  Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r sequence}
#1. 
seq(1,100,4)
Sequence1 <- seq(1,100,4)

#I used the sequence function to create a sequence from 1 - 100, then by adding a 4 to the third comma the sequence goes up by 4. I then assigned the function to an object 'Sequence 1'
#2. 
mean(Sequence1)
median(Sequence1)
# I used the 'mean' and 'median' commands to find the average and median of the object 'Sequence1' that I created in the previous step.
#
#3. 
mean (Sequence1) > median(Sequence1)
# I used the 'greater than' tool on the the 'mean' and 'median' commands to determine if the mean was greater than the median. Since they both equal 49, the mean is not greater than the median. 
```

## Basics Day 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r vectors}
#6
vector_names <- c('Dori', 'Megan', 'Emma', 'Kallie')
vector_grades <- c(46,98,97,96)
vector_pass <- c(FALSE,TRUE,TRUE,TRUE)

# for vector_names I created a list of four students, for vector_grades I assigned each student a grade, for vector_pass I determined if the students passed or failed, if their score was under 50 they were assigned 'FALSE' if their score was over 50 they were assigned 'TRUE'
a <- vector_names
b <- vector_grades
c <- vector_pass

# I then assigned each vector a variable a, b, or c, to simplify.
#7
dataframegrades <- data.frame(a,b,c)

# using the data.frame command I plugged each variable into 'dataframegrades'
#8
names(dataframegrades) <- c("Student Names","Grade","Pass/Fail"); print(dataframegrades)
# Using the 'names' command I then assigned each variable previously created to a more descritive name.


```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: Matrices can only function with one data type, datasets can hold more than one kind of data type. In this dataset we use both numeric and text values.

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r function}
Passing_grades <- function(x){
  if (x > 50)
    print('TRUE')
  else
    print('FALSE')

  # I assigned the function to the object 'Passing_grades' and used the 'if' statement to say that if x was over 50 the output for that variable would be TRUE, and used the 'else' statement to say that all other values (under 50) would output FALSE.
}
Passing_grades2 <- function(x)
  ifelse(x > 50, 'TRUE', 'FALSE')
# I used the 'ifelse' statement and assigned the test value to evaluate whether or not the values in the vector were greater than 50, if they are greater than 50 the output was set to TRUE, if they are less than 50, the output is set to FALSE.

Passing_grades2(vector_grades)
#Passing_grades(vector_grades)

# Passing_grades2, with the ifelse statement worked and was able to evaluate the whole vector, the if and else statements failed since there is more than one value in the vector.
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: 'if' and 'else' statements only work for single values, the 'ifelse' statement can evaluate an entire vector to determine if it meets the function criteria.
