---
title: "Rathmell_Dori_A02_CodingBasics.Rmd"
author: "Dori Rathmell"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: inline
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics Day 1

1.  Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r sequence}
#1. 
seq(1,100,4) # Sequence function creates a sequence from 1-100, skips by 4
Sequence1 <- seq(1,100,4) #assigned the sequence a name

#2. 
mean(Sequence1) # mean command finds the average of Sequence1
median(Sequence1) # median command finds the median of Sequence1

#3. 
mean (Sequence1) > median(Sequence1) 
# greater than tool compares size of mean and median of Sequence1
```

## Basics Day 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r vectors}
#6
vector_names <- c('Dori', 'Megan', 'Emma', 'Kallie') #character vector
vector_grades <- c(46,98,97,96) #numeric vector
vector_pass <- c(FALSE,TRUE,TRUE,TRUE) #logical vector
#created vectors for students, their grades, and whether or not they passed

a <- vector_names
b <- vector_grades
c <- vector_pass
#assigned each vector a variable a, b, or c.

#7
dataframegrades <- data.frame(a,b,c) #created a dataframe using variables

#8
names(dataframegrades) <- c("Student Names","Grade","Pass/Fail") 
#assigning each column in the dataframe a name
print(dataframegrades)
#displaying the dataframe



```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: Matrices can only function with one data type, datasets can hold more
than one kind of data type. In this dataset we use both numeric and text values.

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r function}
Passing_grades <- function(x){
  if (x > 50)
    print('TRUE')
  else
    print('FALSE')
#if else statements print TRUE if value is above 50, FALSE if below 50
}
Passing_grades2 <- function(x)
  ifelse(x > 50, 'TRUE', 'FALSE')
#ifelse command tests if x > 50, if yes prints 'TRUE', if no prints 'FALSE'

Passing_grades2(vector_grades)#applies Passing_grades2 function to vector_grades
#Passing_grades(vector_grades) 
#The if and else statements fail since vector_grades has multiple values


```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: 'if' and 'else' statements only work for single values, the 'ifelse' 
statement can evaluate an entire vector to determine if it meets the function 
criteria.
