---
title: "Assignment 5: Data Visualization"
author: "Dori Rathmell"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
#This code chunk will tidy your knit PDF files, wrapping long code lines
#For it to work, the "formatR" package needs to be installed

#install.packages('formatR')
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)
```

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

The completed exercise is due on Friday, Oct 14th @ 5:00pm.



## Set up your session 
1. Set up your session. Verify your working directory and load the tidyverse, 
lubridate, & cowplot packages. Upload the NTL-LTER processed data files for  
nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy 
[`NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv`] version) and 
the processed data file for the Niwot Ridge litter dataset (use the
[`NEON_NIWO_Litter_mass_trap_Processed.csv`] version). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r}
#1 
getwd()
library(tidyverse)
library(lubridate)
#install.packages(cowplot)
library(cowplot)
Litter_Mass_data <- read.csv('../Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv', 
           stringsAsFactors = TRUE)
Chem_nutrients <- 
  read.csv('../Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv', stringsAsFactors = TRUE)
#opening and naming datasets 
#2 
class(Litter_Mass_data$collectdate)
class(Chem_nutrients$sampledate)
#checking the initial class of date columns in both datasets
Litter_Mass_data$collectDate <- 
  as.Date(Litter_Mass_data$collectDate, format = '%Y-%m-%d')
Chem_nutrients$sampledate <- 
  as.Date(Chem_nutrients$sampledate, format = '%Y-%m-%d')
#reclassifying them as "date" classes
class(Litter_Mass_data$collectDate)
class(Chem_nutrients$sampledate)
#double checking that they are now in the "date" class
```
## Define your theme
3. Build a theme and set it as your default theme.

```{r}
#3
plot_theme <- theme_gray(base_size = 12) +
  theme(axis.text = element_text(color = "Gray"), 
        legend.position = "right")
theme_set(plot_theme)
#setting my default theme to the gray theme 


```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best 
practices for data visualization. Ensure your theme, color palettes, axes, 
and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with 
separate aesthetics for Peter and Paul lakes. Add a line of best fit and 
color it black. Adjust your axes to hide extreme values (hint: change the 
limits using `xlim()` and/or `ylim()`).

```{r}
#4
plot1 <- ggplot(Chem_nutrients, aes(y = po4, x = tp_ug)) +
  geom_point(aes(color = lakename))+
  ylim(0, 50)+
  plot_theme+
  geom_smooth(method=lm, se=FALSE, color = 'black')+
  xlab('Phosphate')+
  ylab('Phosphorus')
#making my first scatter plot, setting the colors and theme + line of best fit
print(plot1)
#displaying scatter plot
```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and 
(c) TN, with month as the x axis and lake as a color aesthetic. Then, create 
a cowplot that combines the three graphs. Make sure that only one legend is 
present and that graph axes are aligned. 

Tip: R has a build in variable called `month.abb` that returns a list of months;
see <https://r-lang.com/month-abb-in-r-with-example>

```{r}
#5
#a
boxplot1 <- 
  ggplot(Chem_nutrients, aes(x = factor(month, levels=c(1:12)), y = temperature_C)) +
  geom_boxplot(aes(color = lakename))+
  plot_theme +
  ylab('Temperature')+
  xlab('Month')+
  scale_x_discrete(labels=month.abb[],drop=F) #converting months from numbers to names
#making the first boxplot
print(boxplot1)
#displaying the first boxplot
#b
boxplot2 <-
  ggplot(Chem_nutrients, aes(x = factor(month, levels=c(1:12)), y = tp_ug)) +
  geom_boxplot(aes(color = lakename))+
  plot_theme+
  ylab('Phosphorus')+
  xlab('Month')+
  scale_x_discrete(labels=month.abb[],drop=F)
#making the second box plot
print(boxplot2)
#displaying the second box plot

#c
boxplot3 <-
  ggplot(Chem_nutrients, aes(x = factor(month, levels=c(1:12)), y = tn_ug)) +
  geom_boxplot(aes(color = lakename))+
  plot_theme+
  xlab('Month')+
  ylab('Nitrogen')+
  scale_x_discrete(labels=month.abb[],drop=F)
#making the third box plot
print(boxplot3)
#displaying the third box plot

#d

library(cowplot)
cowplot1 <- plot_grid(
  boxplot1 + theme(legend.position="none", axis.title.x = element_blank()),
  boxplot2 + theme(legend.position="none", axis.title.x = element_blank()),
  boxplot3 + theme(legend.position='bottom'), #making only one legend present
  align = 'vh',
  ncol = 1,nrow = 3, rel_heights = c(1.5, 1.5, 2)#setting the heights and rows
  #combining all the box plots into one cow plot
)
print(cowplot1)
#displaying cowplot

```

Question: What do you observe about the variables of interest over seasons 
and between lakes? 

> Answer: While temperature is consistent between the two lakes, Peter Lake 
has a higher Nitrogen and Phosphorus content throughout the year.

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the 
"Needles" functional group. Plot the dry mass of needle litter by date and 
separate by NLCD class with a color aesthetic. (no need to adjust the name 
of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into 
three facets rather than separated by color. 

```{r}
#6
Litter_Mass_dataNEW <-
  Litter_Mass_data %>% 
  filter(functionalGroup == 'Needles') %>% 
  group_by(dryMass, collectDate, nlcdClass)
#filtering and grouping the data needed for the plot

litter_plot1 <-
  ggplot(Litter_Mass_dataNEW, aes(x= collectDate, y = dryMass))+
  geom_point(aes(color = nlcdClass))+
  xlab('Date')+
  ylab('Mass of Leaf Litter')+
  plot_theme
print(litter_plot1)
#creating the plot grouped by color 
#7

litter_plot2 <-
  ggplot(Litter_Mass_dataNEW, aes(x = collectDate, y = dryMass)) +
  geom_point(size = 1) +
  plot_theme+
  facet_wrap(vars(nlcdClass), nrow = 3)+
  xlab('Year')+
  ylab('Drymass')
print(litter_plot2)
#creating the plot separated by category
```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: I think plot number 7 is more effective, having the dry mass already 
separated by category even without color is more clear than plot 6 where the 
mass is separated by color but all stacked on top of eachother.